### 2016-12-14 first-submission/ada-ftrl-proofs.tex
Rename directory

\begin{proof}[Proof of Lemma~\ref{lemma:recurrence-solution}]
We prove the lemma by induction on $T$.  The base case $T=0$ is trivial. For $T
\ge 1$, we have
$$
\Delta_T
\le \Delta_{T-1} + \min \left\{a_T, \ \frac{a_T^2}{\Delta_{T-1}} \right\}
\le \sqrt{3 \sum_{t=1}^{T-1} a_t^2} + \min \left\{ a_T, \frac{a_T^2}{\sqrt{3 \sum_{t=1}^{T-1} a_t^2}} \right\}
$$
where the second inequality from the induction hypothesis and the fact that the
function $f(x) = x + \min\{a_T, a_T^2/x\}$ is increasing on $[0,\infty)$.  It
remains to prove that
$$
\sqrt{3 \sum_{t=1}^{T-1} a_t^2} + \min \left\{ a_T, \frac{a_T^2}{\sqrt{3 \sum_{t=1}^{T-1} a_t^2}} \right\}
\le  \sqrt{3 \sum_{t=1}^T a_t^2} \; .
$$
If $a_T = 0$, the inequality clearly holds. Otherwise, $a_T$ is positive and we
divide through by it and make substitution $z=\frac{\sqrt{\sum_{t=1}^{T-1}
a_t^2}}{a_T}$, which leads to
$$
z\sqrt{3} + \min\left\{1,\frac{1}{z\sqrt{3}}\right\} \le \sqrt{3 + 3z^2}
$$
which can be easily checked by considering separately the cases $z \in
[0,\frac{1}{\sqrt{3}})$ and $z \in [\frac{1}{\sqrt{3}}, \infty)$.
\end{proof}

>>>>

\begin{proof}[Proof of Lemma~\ref{lemma:recurrence-solution}]
We prove the lemma by induction on $T$.  The base case $T=0$ is trivial.
For $T \ge 1$, let $f(x) = x + \min\{a_T, a_T^2/x\}$. We have
\begin{align*}
\Delta_T
& \le \Delta_{T-1} + \min \left\{a_T, \ \frac{a_T^2}{\Delta_{T-1}} \right\} \\
& = f \left( \Delta_{T-1} \right) \\
& \le f \left( \sqrt{3 \sum_{t=1}^{T-1} a_t^2} \right) \\
& = \sqrt{3 \sum_{t=1}^{T-1} a_t^2} \ + \ \min \left\{ a_T, \ \frac{a_T^2}{\sqrt{3 \sum_{t=1}^{T-1} a_t^2}} \right\}  \; ,
\end{align*}
where the second inequality follows from the inductive hypothesis $\Delta_{T-1} \le
\sqrt{3 \sum_{t=1}^{T-1} a_t^2}$ and that $f(x)$ is non-positive on $[-\infty,
0)$ and non-negative increasing on $[0,\infty)$. It remains to prove that
$$
\sqrt{3 \sum_{t=1}^{T-1} a_t^2} \ + \ \min \left\{ a_T, \ \frac{a_T^2}{\sqrt{3 \sum_{t=1}^{T-1} a_t^2}} \right\}
\le  \sqrt{3 \sum_{t=1}^T a_t^2} \; .
$$
If $a_T = 0$, the inequality clearly holds. Otherwise, $a_T$ is positive and we
divide through by it and make substitution $z=\frac{\sqrt{\sum_{t=1}^{T-1}
a_t^2}}{a_T}$, which leads to
$$
z\sqrt{3} + \min\left\{1, \frac{1}{z\sqrt{3}}\right\} \le \sqrt{3 + 3z^2}
$$
which can be easily checked by considering separately the cases $z \in
[0,\frac{1}{\sqrt{3}})$ and $z \in [\frac{1}{\sqrt{3}}, \infty)$.
\end{proof}

### 2016-12-14 first-submission/introduction.tex
Rename directory

Majority of existing algorithms for OLO are based on two generic algorithms:
\textsc{Follow The Regularizer Leader} (\textsc{FTRL}) and \textsc{Mirror
Descent} (\textsc{MD}). \textsc{FTRL} dates back to the potential-based
forecaster in \cite[Chapter~11]{Cesa-Bianchi-Lugosi-2006} and its theory was
developed in \cite{Shalev-Shwartz-2007}.  The name \textsc{Follow The
Regularized Leader} comes from~\cite{Abernethy-Hazan-Rakhlin-2008}.
Independently, the same algorithm was proposed in~\cite{Nestorov-2009} for
convex optimization under the name \textsc{Dual Averaging} and later
rediscovered in~\cite{Xiao-2010} for online convex optimization. The general
version with time-varying regularizers was analyzed in
\cite{Orabona-Crammer-Cesa-Bianchi-2014}. \textsc{MD} was originally proposed
in \cite{Nemirovski-Yudin-1983} and later analyzed in~\cite{Beck-Teboulle-2003}
for convex optimization. In the online learning literature it makes the first
appearance, under a different name, in \cite{Kivinen-Warmuth-1997}. A good
description of \textsc{MD} is given in~\cite{Rakhlin-Sridharan-2009}.

Both \textsc{FTRL} and \textsc{MD} are parametrized by a function called a
\emph{regularizer}. The function is assumed to be strongly convex with respect
to a given norm. Based on different regularizers different algorithms with
different properties can be instantiated. A summary of algorithms for OLO is
presented in Table~\ref{table:results}.  All of them are instances of
\textsc{FTRL} or \textsc{MD}.

@@

Only one scale-free algorithm based on \textsc{FTRL} was known. It is the
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} algorithm for
learning with expert advice, where the decision set is bounded.
Algorithms based on \textsc{FTRL} that are ``almost'' scale-free include
\textsc{FTRL-Proximal}~\cite{McMahan-Streeter-2010,McMahan-2014} and
\textsc{AdaGrad FTRL}~\cite{Duchi-Hazan-Singer-2011}.  These algorithms fail to
be scale-free due to ``off-by-one'' issue; see~\cite{McMahan-2014} and also our
discussion in Section~\ref{section:solo-ftrl}.

@@

We design and analyze three scale-free algorithms: \textsc{AdaFTRL},
\textsc{SOLO FTRL} and \textsc{Scale-Free MD}.  \textsc{AdaFTRL} and
\textsc{SOLO FTRL} are based on \textsc{FTRL}.  \textsc{AdaFTRL} is a
generalization of
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} to arbitrary
strongly convex regularizers.  \textsc{SOLO FTRL} can be viewed as the
``correct'' scale-free version of \textsc{AdaGrad
FTRL}~\cite{Duchi-Hazan-Singer-2011} generalized to arbitrary strongly convex
regularizers.  \textsc{Scale-Free MD} is based on \textsc{MD}. It is a
generalization of \textsc{AdaGrad MD}~\cite{Duchi-Hazan-Singer-2011} to
arbitrary strongly convex regularizers, similar to \textsc{Optimistic
MD}~\cite{Rakhlin-Sridharan-2013}.  The three algorithms are presented and
analyzed in Sections~\ref{section:ada-ftrl}, \ref{section:solo-ftrl} and
\ref{section:mirror-descent} respectively.

We prove that the regret of \textsc{AdaFTRL} and \textsc{SOLO FTRL} after $T$
rounds is bounded by $O \left(\sqrt{\sup_{v \in K} f(v)
\sum_{t=1}^T\|\ell_t\|_*^2} \right)$ where $f$ is a non-negative regularizer
that is $1$-strongly convex with respect to a norm $\|\cdot\|$ and
$\|\cdot\|_*$ is its dual norm. For \textsc{Scale-Free MD}, we prove $O
\left(\sqrt{\sup_{u,v \in K} B_f(u,v) \sum_{t=1}^T\|\ell_t\|_*^2} \right)$
where $B_f$ is the Bregman divergence associated with a $1$-strongly convex
regularizer $f$.  As we discuss, in Section~\ref{section:mirror-descent}, the
upper bound for \textsc{Scale-Free MD} is never better than the bound for bound
for \textsc{AdaFTRL} and \textsc{SOLO FTRL}.  In
Section~\ref{section:lower-bound}, we show that the $\sqrt{\sum_{t=1}^T
\|\ell_t\|_*^2}$ term in the bounds is necessary by proving $\frac{D}{\sqrt{8}}
\sqrt{\sum_{t=1}^T\|\ell_t\|_*^2}$ lower bound on the regret of any algorithm
for OLO for any decision set with diameter $D$ with respect to the primal norm
$\|\cdot\|$.

For \textsc{SOLO FTRL}, we prove that regret against a competitor $u \in K$ is
at most $O \left(f(u) \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} + \max_{t=1,2,\dots,T}
\|\ell_t\|_* \sqrt{T} \right)$.  Same as before, $f$ is a non-negative
$1$-strongly convex regularizer. This bound is non-trivial for any decision
set, bounded or unbounded.  The result makes \textsc{SOLO FTRL} the
\textbf{first adaptive algorithm for unbounded decision sets} with a
non-trivial regret bound.

>>>>

The majority of the existing algorithms for OLO are based on two generic algorithms:
\textsc{Follow The Regularizer Leader} (\textsc{FTRL}) and \textsc{Mirror
Descent} (\textsc{MD}). \textsc{FTRL} dates back to the potential-based
forecaster in \cite[Chapter~11]{Cesa-Bianchi-Lugosi-2006} and its theory was
developed in \cite{Shalev-Shwartz-2007}.  The name \textsc{Follow The
Regularized Leader} comes from~\cite{Abernethy-Hazan-Rakhlin-2008}.
Independently, the same algorithm was proposed in~\cite{Nestorov-2009} for
convex optimization under the name \textsc{Dual Averaging} and rediscovered
in~\cite{Xiao-2010} for online convex optimization. The general
version with time-varying regularizers was analyzed in
\cite{Orabona-Crammer-Cesa-Bianchi-2014}. \textsc{MD} was originally proposed
in \cite{Nemirovski-Yudin-1983} and later analyzed in~\cite{Beck-Teboulle-2003}
for convex optimization. In the online learning literature it makes his first
appearance, with a different name, in~\cite{Kivinen-Warmuth-1997}.
%A good description of \textsc{MD} is given in~\cite{Rakhlin-Sridharan-2009}.

Both \textsc{FTRL} and \textsc{MD} are parametrized by a function called 
\emph{regularizer}. The function is assumed to be strongly convex with respect
to a given norm. Based on different regularizers different algorithms with
different properties can be instantiated. A summary of algorithms for OLO is
presented in Table~\ref{table:results}.  All of them are instances of
\textsc{FTRL} or \textsc{MD}.

@@

Only one scale-free algorithm based on \textsc{FTRL} was known. It is the
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} algorithm for
learning with expert advice, where the decision set is bounded.
Algorithms based on \textsc{FTRL} that are ``almost'' scale-free include
\textsc{FTRL-Proximal}~\cite{McMahan-Streeter-2010,McMahan-2014} and
\textsc{AdaGrad FTRL}~\cite{Duchi-Hazan-Singer-2011}.  These algorithms fail to
be scale-free due to ``off-by-one'' issue; see~\cite{McMahan-2014} and the
discussion in Section~\ref{section:solo-ftrl}.

@@

We design and analyze three scale-free algorithms: \textsc{AdaFTRL},
\textsc{SOLO FTRL} and \textsc{Scale-Free MD}.  \textsc{AdaFTRL} and
\textsc{SOLO FTRL} are based on \textsc{FTRL}.  \textsc{AdaFTRL} is a
generalization of
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} to arbitrary
strongly convex regularizers.  \textsc{SOLO FTRL} can be viewed as the
``correct'' scale-free version of \textsc{AdaGrad
FTRL}~\cite{Duchi-Hazan-Singer-2011} generalized to arbitrary strongly convex
regularizers.  \textsc{Scale-Free MD} is based on \textsc{MD}. It is a
generalization of \textsc{AdaGrad MD}~\cite{Duchi-Hazan-Singer-2011} to
arbitrary strongly convex regularizers, similar to \textsc{Optimistic
MD}~\cite{Rakhlin-Sridharan-2013}.  The three algorithms are presented in
Sections~\ref{section:ada-ftrl}, \ref{section:solo-ftrl} and
\ref{section:mirror-descent} respectively.

We prove that the regret of \textsc{AdaFTRL} and \textsc{SOLO FTRL} after $T$
rounds is bounded by $O (\sqrt{\sup_{v \in K} f(v)
\sum_{t=1}^T\|\ell_t\|_*^2} )$ where $f$ is a non-negative regularizer
that is $1$-strongly convex with respect to a norm $\|\cdot\|$ and
$\|\cdot\|_*$ is its dual norm. For \textsc{Scale-Free MD}, we prove $O
(\sqrt{\sup_{u,v \in K} B_f(u,v) \sum_{t=1}^T\|\ell_t\|_*^2} )$
where $B_f$ is the Bregman divergence associated with a $1$-strongly convex
regularizer $f$.  As we discuss, in Section~\ref{section:mirror-descent}, the
upper bound for \textsc{Scale-Free MD} is never better than the bound for bound
for \textsc{AdaFTRL} and \textsc{SOLO FTRL}.  In
Section~\ref{section:lower-bound}, we show that the $\sqrt{\sum_{t=1}^T
\|\ell_t\|_*^2}$ term in the bounds is necessary by proving $\frac{D}{\sqrt{8}}
\sqrt{\sum_{t=1}^T\|\ell_t\|_*^2}$ lower bound on the regret of any algorithm
for OLO for any decision set with diameter $D$ with respect to the primal norm
$\|\cdot\|$.

For \textsc{SOLO FTRL}, we prove that regret against a competitor $u \in K$ is
at most $O (f(u) \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} + \max_{t=1,2,\dots,T}
\|\ell_t\|_* \sqrt{T} )$.  Same as before, $f$ is a non-negative
$1$-strongly convex regularizer. This bound is non-trivial for any decision
set, bounded or unbounded.  The result makes \textsc{SOLO FTRL} the
\textbf{first adaptive algorithm for unbounded decision sets} with a
non-trivial regret bound.
